% The Explanation Impossibility: A Universal Trilemma for Interpreting
% Underspecified Systems
%
% NeurIPS 2026 submission — pages 1–5 (first half)

\documentclass{article}
\usepackage[preprint]{neurips_2026}
% natbib loaded by neurips_2026.sty
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{enumitem}

\DeclareMathOperator{\Var}{Var}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\title{The Explanation Impossibility:\\A Universal Trilemma for Interpreting Underspecified Systems}

\author{
  Anonymous Authors
}

\begin{document}

\maketitle

%% ===================================================================
%% ABSTRACT
%% ===================================================================
\begin{abstract}
We prove that no explanation of an underspecified system can be simultaneously
\emph{faithful} (consistent with the system's native explanation),
\emph{stable} (invariant across equivalent configurations), and
\emph{decisive} (preserving genuine disagreements).
For binary explanation problems---including mechanistic interpretability and
SHAP sign attribution---the impossibility strengthens to a \emph{bilemma}:
faithfulness and stability alone are jointly unachievable.
The result is universal: a single four-line proof, with zero domain-specific
axioms, covers every instance we tested.
We classify which problems admit resolutions and which do not:
the algebraic structure of the explanation space (presence or absence of a
neutral element) determines the ceiling.
We instantiate the framework across mechanistic interpretability (circuit
identification), feature attribution (SHAP rankings), causal discovery
(edge orientation), and attention-based token importance, showing each is a
special case of the same theorem.
When a resolution exists, we characterize the optimal one via the
Hunt--Stein theorem: the Pitman estimator (group-orbit average).
The entire framework---357 theorems from 6 axioms, 0 \texttt{sorry}---is formalized in
Lean~4, with all binary explanation types (\texttt{SHAPSign},
\texttt{FeatureStatus}, \texttt{CounterfactualDir}) defined as
constructive inductive types and the mechanistic interpretability
Rashomon property \emph{derived} by computation (zero axioms).
\end{abstract}

%% ===================================================================
%% 1. INTRODUCTION
%% ===================================================================
\section{Introduction}
\label{sec:intro}

The search for circuits in neural networks is a cornerstone of AI safety.
Mechanistic interpretability aims to identify which internal components---neurons,
attention heads, subnetworks---implement a given behavior, with the ultimate goal
of verifying that a model's reasoning is aligned with human intent.
We prove that this search faces a fundamental mathematical constraint:
no circuit explanation can be simultaneously faithful to the network's internal
structure, stable across functionally equivalent networks, and decisive about
which decomposition is correct.
For binary circuit questions---``is this component part of the circuit or
not?''---the situation is strictly worse: faithfulness and stability alone
are jointly impossible, regardless of whether decisiveness is required.

This result is not specific to mechanistic interpretability.
The same impossibility governs every explanation method we tested:
SHAP feature rankings, causal edge orientations, attention-based token
importance, feature selection, and counterfactual directions.
All are instances of a single abstract theorem that requires
\emph{zero domain-specific axioms}---only the observation that underspecified
systems admit equivalent configurations with incompatible native explanations
(the Rashomon property).

The impossibility admits resolution for some problems but not others.
The key structural diagnostic is whether the explanation space contains a
\emph{neutral element}---a value compatible with every native explanation.
Feature rankings admit ties (a neutral element), so ensemble averaging
(DASH) resolves the trilemma.
Circuit decompositions do not admit a neutral element, so the bilemma is
inescapable; the only resolution is to enrich the explanation space by
reporting equivalence classes of circuits.
This classification is complete: the presence or absence of neutral and
committal elements determines exactly which pairs of axioms can be jointly
satisfied.

\paragraph{Contributions.}
\begin{enumerate}[leftmargin=*,label=(\arabic*)]
\item \textbf{Abstract framework and universal impossibility.}
  We define an \texttt{ExplanationSystem} $(\Theta, H, Y, \mathrm{observe},
  \mathrm{explain}, \mathrm{incompatible})$ and prove that no explanation
  $E : \Theta \to H$ of a system with the Rashomon property can be
  faithful, stable, and decisive (Theorem~\ref{thm:trilemma}).
  The proof is four lines and uses zero domain-specific axioms.

\item \textbf{The bilemma.}
  For maximally incompatible explanation spaces (compatible $\Rightarrow$
  equal), faithful + stable alone is impossible
  (Theorem~\ref{thm:bilemma})---strictly stronger than the trilemma.
  This applies to all binary explanation problems: SHAP sign, circuit
  decomposition, feature selection, and counterfactual direction
  (increase vs.\ decrease a feature).
  Each is formalized as a named bilemma instance
  (\texttt{shap\_sign\_bilemma}, \texttt{feature\_selection\_bilemma},
  \texttt{counterfactual\_bilemma}).

\item \textbf{Tightness classification.}
  We give a complete $2 \times 2$ classification
  (Table~\ref{tab:tightness}) showing which pairs of axioms are jointly
  achievable as a function of the explanation space's algebraic structure
  (neutral elements $\times$ committal elements).

\item \textbf{Instances.}
  We instantiate the framework across mechanistic interpretability
  (\S\ref{sec:mech-interp}), feature attribution (\S\ref{sec:attribution}),
  causal discovery (\S\ref{sec:causal}), and attention maps
  (\S\ref{sec:attention}), with additional instances in the supplement.

\item \textbf{Optimal resolution via Hunt--Stein.}
  When a resolution exists (neutral element present), the optimal
  explanation is the Pitman estimator---the average over the symmetry group
  orbit. For feature rankings, this is DASH (ensemble SHAP averaging).
  For causal discovery, this is the CPDAG.

\item \textbf{Lean~4 formalization.}
  The full framework is formalized across 58 files containing 357
  theorems and lemmas with 0~\texttt{sorry} and only 6 axioms
  (all infrastructure: type declarations and measure setup).
  The mechanistic interpretability Rashomon property is \emph{derived}
  from constructive finite types, not axiomatized.
  Three domain-specific inductive types---\texttt{SHAPSign},
  \texttt{FeatureStatus}, \texttt{CounterfactualDir}---are
  defined constructively with zero axioms.
\end{enumerate}

\paragraph{Scope separation.}
A companion paper~\citep{anonymous2026attribution} develops the
quantitative theory for feature rankings under collinearity: the
$1/(1{-}\rho^2)$ divergence rate, Spearman instability bounds, and
GBDT/Lasso/neural-net instantiations.
The present paper abstracts the core mechanism into a domain-independent
framework, proves the bilemma (absent from the companion),
and extends the impossibility to mechanistic interpretability,
causal discovery, attention maps, and counterfactual directions.

%% ===================================================================
%% 2. THE FRAMEWORK
%% ===================================================================
\section{The Framework}
\label{sec:framework}

\begin{definition}[Explanation System]
\label{def:explanation-system}
An \emph{explanation system} is a tuple $S = (\Theta, H, Y,
\mathrm{observe}, \mathrm{explain}, \mathrm{incompatible})$ where:
\begin{itemize}[leftmargin=*]
\item $\Theta$ is the \emph{configuration space} (e.g., trained models,
  network weight vectors, DAGs);
\item $H$ is the \emph{explanation space} (e.g., feature rankings, circuit
  decompositions, edge orientations);
\item $Y$ is the \emph{observable space} (e.g., predictions, I/O functions,
  conditional independence structures);
\item $\mathrm{observe} : \Theta \to Y$ maps each configuration to its
  observable behavior;
\item $\mathrm{explain} : \Theta \to H$ maps each configuration to its
  \emph{native} explanation (the ground-truth interpretation that the
  configuration's internals warrant);
\item $\mathrm{incompatible} : H \times H \to \mathrm{Prop}$ is an
  irreflexive relation on explanations capturing genuine disagreement.
\end{itemize}
\end{definition}

An \emph{explanation method} is any function $E : \Theta \to H$.
We require three properties:

\paragraph{Faithfulness.}
$E$ is \emph{faithful} if it never contradicts the system's native
explanation:
\[
  \forall\, \theta \in \Theta,\quad
  \neg\,\mathrm{incompatible}(E(\theta),\; \mathrm{explain}(\theta)).
\]
This is deliberately weak: $E(\theta)$ need not \emph{equal}
$\mathrm{explain}(\theta)$, only be compatible with it.
A faithful explanation may be vague (e.g., report a tie) but must not
actively contradict the model.

\paragraph{Stability.}
$E$ is \emph{stable} if equivalent configurations receive the same
explanation:
\[
  \forall\, \theta_1, \theta_2 \in \Theta,\quad
  \mathrm{observe}(\theta_1) = \mathrm{observe}(\theta_2)
  \;\Longrightarrow\;
  E(\theta_1) = E(\theta_2).
\]
This is the explanation analogue of the statistical principle that
functionally equivalent models should yield the same interpretation.
Without stability, explanations depend on arbitrary implementation details
(random seed, initialization, training order) rather than the system's
behavior.

\paragraph{Decisiveness.}
$E$ is \emph{decisive} if it preserves the incompatibility of native
explanations:
\[
  \forall\, \theta_1, \theta_2 \in \Theta,\quad
  \mathrm{incompatible}(\mathrm{explain}(\theta_1),\;
  \mathrm{explain}(\theta_2))
  \;\Longrightarrow\;
  \mathrm{incompatible}(E(\theta_1),\; E(\theta_2)).
\]
A decisive method does not smooth over genuine disagreements between
configurations.
When two models truly disagree about which feature is more important
(or which circuit implements a behavior), a decisive method reports
that disagreement rather than collapsing it.

\paragraph{The Rashomon property.}
An explanation system has the \emph{Rashomon property} if there exist
configurations $\theta_1, \theta_2 \in \Theta$ such that
$\mathrm{observe}(\theta_1) = \mathrm{observe}(\theta_2)$ and
$\mathrm{incompatible}(\mathrm{explain}(\theta_1),\;
\mathrm{explain}(\theta_2))$.
That is, the system admits equivalent configurations whose native
explanations genuinely disagree.
This is the mechanism behind model multiplicity: the same predictive
performance can arise from models with contradictory internal structures.

%% ===================================================================
%% 3. THE IMPOSSIBILITY
%% ===================================================================
\section{The Impossibility}
\label{sec:impossibility}

\begin{theorem}[The Explanation Trilemma]
\label{thm:trilemma}
Let $S = (\Theta, H, Y, \mathrm{observe}, \mathrm{explain},
\mathrm{incompatible})$ be an explanation system where
$\mathrm{incompatible}$ is irreflexive.
If $S$ has the Rashomon property, then no explanation method
$E : \Theta \to H$ can be simultaneously faithful, stable, and decisive.
\end{theorem}

\begin{proof}
Let $\theta_1, \theta_2$ be Rashomon witnesses:
$\mathrm{observe}(\theta_1) = \mathrm{observe}(\theta_2)$ and
$\mathrm{incompatible}(\mathrm{explain}(\theta_1),
\mathrm{explain}(\theta_2))$.
(1)~Decisive: $\mathrm{incompatible}(E(\theta_1), E(\theta_2))$.
(2)~Stable: $E(\theta_1) = E(\theta_2)$.
(3)~Substituting: $\mathrm{incompatible}(E(\theta_1), E(\theta_1))$.
(4)~This contradicts irreflexivity.
\end{proof}

\begin{theorem}[Tightness]
\label{thm:tightness}
The trilemma is tight: each pair of axioms is jointly satisfiable.
\end{theorem}
\begin{proof}[Proof sketch]
(F+S, drop D): Let $E$ be a constant function mapping every $\theta$ to a
  neutral element $h_0$ compatible with all native explanations. Then $E$ is
  trivially faithful and stable, but not decisive.
(F+D, drop S): Let $E = \mathrm{explain}$. Then $E$ is faithful and
  decisive by definition, but not stable (Rashomon witnesses get different
  explanations).
(S+D, drop F): Let $E$ be a constant function mapping every $\theta$ to
  some fixed $h^*$ that is incompatible with at least one native
  explanation. Then $E$ is stable and (vacuously) decisive on pairs that
  agree, but unfaithful.
\end{proof}

\begin{theorem}[The Bilemma]
\label{thm:bilemma}
Let $S$ be an explanation system with the Rashomon property, where $H$ is
\emph{maximally incompatible}: $\neg\,\mathrm{incompatible}(h_1, h_2)
\Rightarrow h_1 = h_2$.
Then no explanation method $E : \Theta \to H$ can be simultaneously
faithful and stable.
\end{theorem}
\begin{proof}
Since incompatibility is the complement of equality, faithfulness
($\neg\,\mathrm{incompatible}(E(\theta), \mathrm{explain}(\theta))$)
forces $E(\theta) = \mathrm{explain}(\theta)$ for all $\theta$.
But then $E = \mathrm{explain}$ is decisive (it inherits the
incompatibility structure of $\mathrm{explain}$).
So $E$ is faithful, stable, and decisive, contradicting
Theorem~\ref{thm:trilemma}.
\end{proof}

\paragraph{The tightness classification.}
Whether faithful + stable is achievable depends on the algebraic structure
of the explanation space $H$.
A \emph{neutral element} $h_0 \in H$ satisfies
$\neg\,\mathrm{incompatible}(h_0, h)$ for all $h \in H$ (it is compatible
with every native explanation).
A \emph{committal element} $h^* \in H$ satisfies
$\mathrm{incompatible}(h^*, h)$ for some $h \in H$ (it actively disagrees
with at least one native explanation).

\begin{table}[t]
\centering
\caption{Tightness classification. Which pairs of axioms are jointly
achievable depends on whether $H$ contains neutral and committal
elements. F+S+D is never achievable (the trilemma).}
\label{tab:tightness}
\begin{tabular}{@{}lcc@{}}
\toprule
& Neutral exists & No neutral \\
\midrule
\textbf{Committal exists} & F+D,\; F+S,\; S+D & F+D,\; S+D \\
\textbf{No committal}     & F+D,\; F+S         & F+D only \\
\bottomrule
\end{tabular}
\end{table}

The key insight: \emph{the explanation space's algebraic structure
determines the ceiling.}
Feature rankings have ties (neutral elements), so DASH achieves F+S.
Circuit decompositions have no neutral element (every decomposition commits
to a structure), so F+S is impossible---the bilemma is inescapable.
This yields a \emph{direction theorem}: coverage conflict (the absence of neutral elements) is
anti-monotone in $|H|$---enriching $H$ with neutral elements weakens the
impossibility, the opposite of Arrow's theorem where enlarging the alternative set
strengthens the impossibility by enabling preference cycles.
The diagnostic is actionable: before building an explanation method,
check whether $H$ has a neutral element.
If not, no method can be both faithful and stable, and the only resolution
is to enrich $H$ (e.g., report equivalence classes instead of individual
explanations).

\paragraph{Coverage conflict diagnostic.}
We formalize the diagnostic as a predicate \texttt{hasCoverageConflict}
(\texttt{BeyondBinary.lean}):
an explanation system has \emph{coverage conflict} if every element
$h \in H$ is incompatible with some native explanation value
($\forall\, h,\; \exists\, \theta,\;
\mathrm{incompatible}(h, \mathrm{explain}(\theta))$).
Coverage conflict implies no neutral element exists
(Lean-verified: \texttt{coverageConflict\_implies\_no\_neutral}),
so the bilemma applies.
Conversely, a neutral element destroys coverage conflict
(\texttt{neutral\_destroys\_coverageConflict}).
The tightness dichotomy---F+S achievable $\Leftrightarrow$ absence
of coverage conflict---is proved as a single theorem
(\texttt{tightness\_dichotomy}).

%% ===================================================================
%% 4. INSTANCES
%% ===================================================================
\section{Instances}
\label{sec:instances}

We instantiate the framework across four explanation domains.
Each requires only specifying the tuple $(\Theta, H, Y, \mathrm{observe},
\mathrm{explain}, \mathrm{incompatible})$ and verifying the Rashomon
property; the impossibility then follows from Theorem~\ref{thm:trilemma}
or~\ref{thm:bilemma} with no additional proof.

%% ----- Instance 1: Mechanistic Interpretability -----
\subsection{Instance 1: Mechanistic Interpretability}
\label{sec:mech-interp}

\paragraph{Setup.}
$\Theta$ is the space of neural network configurations (weight vectors);
$H$ is the space of circuit decompositions (which subnetwork implements a
given behavior);
$Y$ is the space of input-output functions.
Two networks are equivalent ($\mathrm{observe}(\theta_1) =
\mathrm{observe}(\theta_2)$) if they compute the same function.
Two decompositions are incompatible if they identify different components
as implementing the behavior.

\paragraph{The bilemma applies.}
Circuit decomposition is a binary question at the component level:
each neuron or attention head is either part of the circuit or not.
The explanation space is maximally incompatible
($\neg\,\mathrm{incompatible}(d_1, d_2) \Rightarrow d_1 = d_2$),
so Theorem~\ref{thm:bilemma} applies directly: no circuit explanation
can be simultaneously faithful and stable, regardless of whether
decisiveness is required.

\paragraph{Rashomon is derived.}
In the Lean formalization, the Rashomon property for circuits is
\emph{derived} from constructive finite types, not axiomatized.
We define two configurations (\texttt{circuitAlpha},
\texttt{circuitBeta}) computing the same function with different
decompositions; the Lean kernel verifies
$\mathrm{observe}(\texttt{circuitAlpha}) =
\mathrm{observe}(\texttt{circuitBeta})$ and
$\mathrm{incompatible}(\mathrm{explain}(\texttt{circuitAlpha}),
\mathrm{explain}(\texttt{circuitBeta}))$
by \texttt{decide} (zero axioms).

\paragraph{Empirical grounding.}
\citet{meloux2025circuits} found 85 distinct valid circuits with zero
circuit error for a simple XOR task, each admitting an average of 535.8
valid interpretations.
Less than 2\% of networks had a unique circuit mapping.
This extreme multiplicity confirms that the Rashomon property holds with
overwhelming force for circuit explanations in practice.

\paragraph{Resolution.}
The bilemma forecloses averaging-based resolutions because the circuit
explanation space has no neutral element.
The resolution requires \emph{enriching} the explanation space: report the
equivalence class of valid circuits---the computational structure shared
across \emph{all} valid decompositions---rather than any individual
decomposition.
This is the circuit analogue of the CPDAG in causal discovery: report what
is shared, acknowledge what is underdetermined.

\paragraph{Implication for AI safety.}
Safety cases that depend on identifying ``the circuit'' for a behavior are
subject to this impossibility: a circuit found in one training run may not
exist in a functionally equivalent run.

%% ----- Instance 2: Feature Attribution -----
\subsection{Instance 2: Feature Attribution (SHAP)}
\label{sec:attribution}

\paragraph{Setup.}
$\Theta$ is the space of trained models (e.g., GBDTs with different
random seeds);
$H$ is the space of feature importance rankings;
$Y$ is the space of predictions (loss-equivalent models map to the same
$y$).
Two rankings are incompatible if they order some feature pair in opposite
directions (feature~$j$ ranked above feature~$k$ vs.\ below).

\paragraph{The trilemma applies (not the bilemma).}
The ranking space admits \emph{ties} as a neutral element: a ranking that
assigns equal importance to features $j$ and $k$ is compatible with both
$j > k$ and $j < k$.
Because a neutral element exists, F+S is achievable (top-left cell of
Table~\ref{tab:tightness})---the bilemma does not apply.
However, ties sacrifice decisiveness, so the full trilemma still holds.

\paragraph{Resolution: DASH.}
DASH (Deterministic Averaged SHAP) averages SHAP values across an ensemble
of equivalent models, producing ties for feature pairs where the ensemble
disagrees.
These ties are the neutral elements that make F+S achievable.
DASH is provably optimal: by the Hunt--Stein theorem for the permutation
group $S_m$ acting on within-group features, the best equivariant estimator
is the Pitman estimator---the group-orbit average---which is exactly DASH.
Empirically, DASH resolves the instability on 68\% of 77 OpenML datasets
tested (under the standard workflow; the model-only component is ${\sim}63\%$
of the total variation).
The full quantitative treatment---$1/(1{-}\rho^2)$ divergence
rate, Spearman bounds, and model-specific
instantiations---appears in~\citet{anonymous2026attribution}.

%% ----- Instance 3: Causal Discovery -----
\subsection{Instance 3: Causal Discovery}
\label{sec:causal}

\paragraph{Setup.}
$\Theta$ is the space of DAGs in a Markov equivalence class;
$H$ is the space of fully oriented edge sets;
$Y$ is the conditional independence (CI) structure shared by all DAGs in
the class.
Two edge orientations are incompatible if they disagree on the direction of
at least one edge.

\paragraph{The Rashomon property.}
Markov equivalence is the canonical example of observational equivalence:
DAGs that encode the same CI structure are indistinguishable from
observational data.
For the minimal case (3-node chain $A \to B \to C$ vs.\ fork
$A \leftarrow B \to C$), same CI structure, different edge orientations.
In the Lean formalization, this is derived by \texttt{decide} (zero
axioms).

\paragraph{Impossibility.}
No stable orientation rule can be faithful to all datasets when the
equivalence class contains $\geq 2$ DAGs.
``Stable'' means the rule gives the same answer for all datasets consistent
with the CI structure; ``faithful'' means the answer matches the
data-optimal DAG.
Since different datasets can make different DAGs optimal, stability forces
a single answer, which cannot match all data-optimal DAGs simultaneously.

\paragraph{Resolution: CPDAG.}
The standard resolution in causal discovery is the CPDAG (completed
partially directed acyclic graph): orient edges that have the same
direction in all Markov-equivalent DAGs, leave the rest undirected.
This is precisely the neutral-element resolution: undirected edges are
compatible with both orientations.
The CPDAG achieves F+S by sacrificing decisiveness on the ambiguous edges.

%% ----- Instance 4: Attention Maps -----
\subsection{Instance 4: Attention Maps}
\label{sec:attention}

\paragraph{Setup.}
$\Theta$ is the space of transformer configurations (weight vectors after
different training runs);
$H$ is the space of token importance scores (or rankings) derived from
attention weights;
$Y$ is the space of input-output mappings.
\citet{damour2022underspecification} established that equivalent networks
(same loss on held-out data) can have markedly different internal
representations, including attention patterns.

\paragraph{The bilemma applies for argmax attention.}
When the explanation is the identity of the most-attended token
($H = \{\text{token}_1, \ldots, \text{token}_n\}$ with
$\mathrm{incompatible}(t_i, t_j) \Leftrightarrow i \neq j$),
the explanation space is maximally incompatible.
Theorem~\ref{thm:bilemma} applies: no method for identifying ``the most
important token'' can be simultaneously faithful and stable.

\paragraph{The trilemma applies for full attention distributions.}
When $H$ is the space of full attention distributions (probability vectors
over tokens), ties exist as neutral elements (uniform attention is
compatible with any native distribution under a suitable incompatibility
relation).
The trilemma applies but the bilemma does not; F+S is achievable via
averaging.

\paragraph{Empirical evidence.}
Using DistilBERT with 10 weight-perturbed variants ($\sigma = 0.02$),
88\% of adjacent token pairs exhibit flip rates exceeding 10\%.
Under actual fine-tuning (5 seeds, 1 epoch on SST-2), 14.5\% of pairs
are unstable---structural instability that persists across training
durations (identical at 1 and 3 epochs).
Mean Spearman correlation of token rankings across variants is 0.637,
comparable to single-model SHAP instability at moderate collinearity
($\rho \approx 0.5$).

% === PAGES 6-10 (merged from neurips_universal_part2.tex) ===
\input{neurips_universal_part2}

\bibliographystyle{plainnat}
\bibliography{references}

\end{document}
