\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsfonts,amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage{booktabs}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{axiom}{Axiom}

\title{Supplementary Material:\\The Attribution Impossibility}
\author{}
\date{}

\begin{document}
\maketitle

\section{Complete Axiom System}

Table~\ref{tab:axioms} lists the 6 property axioms in the formalization
(the core impossibility theorem requires none of them---only the Rashomon
property as hypothesis).
The formalization uses 12 axioms total: 5 type/constant declarations
(\texttt{Model}, \texttt{numTrees}, \texttt{numTrees\_pos}, \texttt{attribution},
\texttt{splitCount}, \texttt{firstMover}), the 5 property axioms below,
the DASH symmetry axiom, and one quantitative Spearman bound.

\begin{table}[h]
\centering
\caption{Complete axiom inventory.}
\label{tab:axioms}
\begin{tabular}{@{}llp{6cm}@{}}
\toprule
\textbf{Axiom} & \textbf{Lean~4 name} & \textbf{Justification} \\
\midrule
First-mover surjectivity & \texttt{firstMover\_surjective} & DGP symmetry: each feature can be first-mover \\
Split count (first-mover) & \texttt{splitCount\_firstMover} & Gaussian conditioning (Lemma~1 in main text) \\
Split count (non-first-mover) & \texttt{splitCount\_nonFirstMover} & Residual signal after first-mover absorbs $\rho$-component \\
Proportionality & \texttt{attribution\_proportional} & Uniform contribution model (Assumption~7) \\
Spearman bound & \texttt{spearman\_classical\_bound} & Combinatorial rank argument: tied non-first-movers \\
Attribution sum symmetry & \texttt{attribution\_sum\_symmetric} & DGP symmetry + balanced ensemble hypothesis \\
\bottomrule
\end{tabular}
\end{table}

\section{Proof Architecture}

The formalization comprises 13 Lean~4 files (12 axioms, 28 substantive theorems):

\begin{verbatim}
Defs.lean (11 axioms, type definitions)
  |
  +-- Trilemma.lean (RashimonProperty, attribution_impossibility)
  |     |
  |     +-- Iterative.lean (IterativeOptimizer, iterative_impossibility)
  |     |     +-- Lasso.lean (lasso_impossibility)
  |     |     +-- NeuralNet.lean (nn_impossibility)
  |     |
  |     +-- General.lean (GBDT instance, gbdt_impossibility)
  |
  +-- SplitGap.lean (split_gap_exact, split_gap_ge_half)
  +-- Ratio.lean (attribution_ratio, ratio_tendsto_atTop)
  +-- SpearmanDef.lean (spearmanCorr definition, derived bound)
  |
  +-- Impossibility.lean (not_equitable, not_stable, impossibility)
  +-- Corollary.lean (consensus_equity, consensus_difference_zero)
  +-- RandomForest.lean (contrast case, documentation only)
\end{verbatim}

\section{Inconsistencies Found During Formalization}

Lean's type checker caught three issues in the initial axiom system:

\begin{enumerate}
\item \textbf{First-mover balance (Axiom~6, original):}
  Originally stated universally over all model arrays---a constant
  model function trivially derived $\bot$ (the first-mover count for
  two different features cannot both equal the total). \emph{Fix:}
  Replaced with an explicit \texttt{IsBalanced} predicate as a
  hypothesis.

\item \textbf{Attribution sum symmetry (Axiom~7, original):}
  Combined with Axioms~2--4, the original unconditional version
  derived $\bot$ for unbalanced ensembles: the summed attributions
  cannot be equal when the first-mover counts differ. \emph{Fix:}
  Conditioned on \texttt{IsBalanced}.

\item \textbf{Split count type:}
  Originally \texttt{splitCount} returned $\mathbb{N}$, but
  $T/(2-\rho^2)$ is generally irrational, creating a type mismatch.
  \emph{Fix:} Changed to $\mathbb{R}$ (idealized leading-order values).
\end{enumerate}

The first two are genuine logical inconsistencies that would have been
difficult to detect by inspection of an informal proof. The third is a
modeling assumption made explicit by the type system.

\section{Gaussian Conditioning Argument}

The split count formulas (Axioms~2--3) are justified by the following
argument. Consider two features $X_j, X_k$ with correlation $\rho$.
At tree $t$, the available signal for feature $k$ after feature $j$
has been selected as root split is:
\[
  X_k \mid X_j = X_k - \rho X_j,
\]
which has variance $1 - \rho^2$. The expected number of splits for the
first-mover sums a geometric series over $T$ trees, yielding
$T/(2 - \rho^2)$. The non-first-mover receives the residual signal,
yielding $(1-\rho^2)T/(2-\rho^2)$. Full derivation with explicit
series computation is in the companion paper.

\section{SymPy Verification}

All algebraic consequences have been independently verified:

\begin{verbatim}
# dash-shap/paper/proofs/verify_lemma6_algebra.py
#
# Verifies:
#   split_gap = rho^2 * T / (2 - rho^2)
#   attribution_ratio = 1 / (1 - rho^2)
#   split_gap >= rho^2 * T / 2  (for rho in (0,1))
#
# Result: ALL CHECKS PASS
\end{verbatim}

\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{remark}[theorem]{Remark}

\section{Extended Results: Universality and Relaxation Characterization}

The results in this section extend the main paper's impossibility theorem
in two directions: (i) showing the Rashomon property is inevitable for
symmetric model classes, making the impossibility inescapable; and
(ii) partially characterizing the remaining two relaxation paths.

\subsection{Rashomon Inevitability Under Symmetry}

The Attribution Impossibility (Theorem~1 in the main text) holds for
any model class satisfying the Rashomon property. We now show this
property is inevitable for symmetric model classes.

\begin{definition}[Permutation closure]
\label{def:perm-closure}
A model class $\mathcal{F}$ is \emph{permutation-closed} within
group $\ell$ if for any $f \in \mathcal{F}$ and any permutation
$\pi$ of features within group $\ell$, the composed model
$f \circ \pi \in \mathcal{F}$.
\end{definition}

This condition is satisfied by any model class without built-in
feature ordering: neural networks (permute input neurons),
gradient-boosted trees (permute split features), Lasso (permute
covariates), and random forests. It fails only for model classes
with hard-coded feature preferences.

\begin{theorem}[Rashomon from symmetry]
\label{thm:rashomon-symmetry}
Let the DGP be symmetric within group $\ell$ (permuting features
$j \leftrightarrow k$ leaves the population loss invariant).
Let $\mathcal{F}$ be permutation-closed within group $\ell$.
Then for any model $f \in \mathcal{F}$ with
$\varphi_j(f) \neq \varphi_k(f)$, the permuted model
$f' = f \circ \pi_{jk}$ satisfies:
\begin{enumerate}
    \item $L(f') = L(f)$ (same population loss),
    \item $\varphi_k(f') = \varphi_j(f)$ and
          $\varphi_j(f') = \varphi_k(f)$ (attributions swap).
\end{enumerate}
In particular, if $\varphi_j(f) > \varphi_k(f)$, then
$\varphi_k(f') > \varphi_j(f')$, and the Rashomon property holds.
\end{theorem}

\begin{proof}
By DGP symmetry, the joint distribution of $(X, Y)$ is invariant
under permutation of features $j$ and $k$ within group $\ell$.
Therefore $L(f') = L(f \circ \pi_{jk}) = L(f)$, since the
population loss depends only on the joint distribution.
By construction, $f'$ applies $f$ to permuted inputs, so its
reliance on feature $k$ equals $f$'s reliance on feature $j$:
$\varphi_k(f') = \varphi_j(f)$ and vice versa.
\end{proof}

\begin{theorem}[Attribution non-degeneracy]
\label{thm:non-degeneracy}
Let $A$ be a stochastic training algorithm with continuous dependence
on its random seed (e.g., SGD with random initialization, bootstrap
sampling, random feature selection). For any DGP with $\rho > 0$ and
$n < \infty$ training samples, $\Pr[\varphi_j(f) \neq \varphi_k(f)] = 1$
for features $j, k$ in the same collinear group.
\end{theorem}

\begin{proof}
With finite samples, the empirical correlation between $X_j$ and $X_k$
is $\hat\rho = \rho + O(1/\sqrt{n})$, which is almost surely irrational.
Under continuous dependence on the data, the training algorithm's feature
utilization varies continuously with $\hat\rho$, so the event
$\{\varphi_j(f) = \varphi_k(f)\}$ has measure zero in the joint
randomness of the data and the algorithm.
\end{proof}

\begin{theorem}[Rashomon inevitability]
\label{thm:rashomon-inevitability}
Let $A$ be a stochastic, symmetric training algorithm
($A$ applied to a within-group-permuted dataset produces the
permuted model in distribution) for a permutation-closed model class
$\mathcal{F}$. For any DGP with $\rho > 0$, the Rashomon property
holds: for any $j, k$ in the same collinear group, there exist
models $f, f'$ in the support of $A$ with $\varphi_j(f) > \varphi_k(f)$
and $\varphi_k(f') > \varphi_j(f')$.
\end{theorem}

\begin{proof}
By Theorem~\ref{thm:non-degeneracy},
$\Pr[\varphi_j(f) \neq \varphi_k(f)] = 1$, so either
$\varphi_j > \varphi_k$ or $\varphi_k > \varphi_j$ almost surely.
By algorithmic symmetry,
$\Pr[\varphi_j > \varphi_k] = \Pr[\varphi_k > \varphi_j]$.
Since these probabilities sum to~1, each equals~$1/2$.
Both events have positive probability, so both are realized.
\end{proof}

\begin{remark}
Theorem~\ref{thm:rashomon-inevitability} makes the Attribution
Impossibility inescapable for any standard ML pipeline under
collinearity. The only escapes are: (i)~$\rho = 0$ (no collinearity),
(ii)~a deterministic model class producing a single model, or
(iii)~an asymmetric algorithm with built-in feature preferences.
\end{remark}

\subsection{The Price of Dropping Faithfulness}

The main text resolves the impossibility by dropping completeness
(DASH: ties for symmetric features). What if we instead drop
faithfulness---allowing a stable, complete ranking that does not
reflect any specific model?

\begin{theorem}[Unfaithfulness lower bound]
\label{thm:unfaithfulness}
Let $j, k$ be symmetric features (same group, same population-level
relationship with $Y$). Any stable, complete ranking $\succ$ has
expected unfaithfulness at least $1/2$ for the pair $(j,k)$:
\[
    \Pr_f\bigl[\, j \succ k \;\text{but}\;
    \varphi_k(f) > \varphi_j(f) \,\bigr] \;=\; \frac{1}{2},
\]
where the probability is over models $f$ drawn from the symmetric
training distribution.
\end{theorem}

\begin{proof}
By DGP symmetry, the distribution of $\varphi_j(f) - \varphi_k(f)$
is symmetric about zero:
$\Pr[\varphi_j(f) > \varphi_k(f)] = \Pr[\varphi_k(f) > \varphi_j(f)] = 1/2$.
Any stable ranking must fix $j \succ k$ or $k \succ j$ independently
of $f$. Whichever it chooses disagrees with exactly half the models.
\end{proof}

\begin{proposition}[Optimal unfaithful ranking]
\label{prop:optimal-unfaithful}
The ranking $j \succ^* k \Leftrightarrow \mathbb{E}[\varphi_j] > \mathbb{E}[\varphi_k]$
minimizes expected unfaithfulness over the symmetric model distribution.
For symmetric features, $\mathbb{E}[\varphi_j] = \mathbb{E}[\varphi_k]$,
so $\succ^*$ assigns a tie---recovering the ``drop completeness'' solution.
\end{proposition}

\begin{proof}
The ranking $\succ^*$ is the Bayes-optimal binary decision for
``is $\varphi_j(f) > \varphi_k(f)$?'' under the model distribution,
minimizing $0$-$1$ loss. For symmetric features the posterior is exactly
$1/2$, so the optimal decision is indeterminate (a tie).
\end{proof}

\begin{theorem}[Relaxation path convergence]
\label{thm:path-convergence}
The ``drop faithfulness'' and ``drop completeness'' relaxation paths
converge to the same solution: population-level attributions with ties
for symmetric features. Consequently, the impossibility trilemma
collapses to a single tradeoff axis parameterized by ensemble size~$M$:
\begin{itemize}
    \item At $M = 1$: faithfulness and completeness, but no stability;
    \item As $M \to \infty$: stability and between-group faithfulness,
          but within-group ties (completeness relaxed).
\end{itemize}
Between-group faithfulness is preserved at all~$M$.
\end{theorem}

\begin{proof}
By Proposition~\ref{prop:optimal-unfaithful}, the optimal stable
complete ranking assigns ties to symmetric features---identical to the
``drop completeness'' solution (DASH). Both paths yield population-level
attributions. The $M$-parameterization follows from DASH consensus:
$\bar\varphi_j = \frac{1}{M}\sum_{i=1}^M \varphi_j(f_i)$, which
interpolates between a single model ($M=1$, faithful but unstable)
and the population mean ($M \to \infty$, stable but tied within groups).
\end{proof}

\end{document}
